---

# @color[#4286f4]@size[3em](Modelling Item Worth Based on Rankings)

@color[#ffffff](Heather Turner)

@color[#ffffff](May 15 2018)


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, fig.path = "../talks-svg/eRum2018/",
               fig.show = "hide", dev = "svg", comment = "#")
makeLink <- function(label, title, alt, width){
    cat('<img src="https://raw.githubusercontent.com/hturner/test/master/',
        label, '-1.svg?sanitize=true" title="', 
        title, '" alt="', alt, '" width="', width, 'px" />', sep = "")
}               
```

---

## Rankings

Rankings arise in a number of settings

 - Finishing order in a set of races
 - Consumer preferences in market research
 
What is the worth of each item?

---

## Luce's Axiom

> Probability of choosing item A over item B unaffected by other items 

Suppose we have a set of $J$ items

$$S = \{i_1, i_2, \ldots, i_J\}$$

Then under Luce's axiom

`$$P(j | S) = \frac{\alpha_{j}}{\sum_{i \in S} \alpha_i}$$`

where `$\alpha_i$` represents the **worth** of item $i$.

---

## Plackett-Luce Model

Consider a ranking of $J$ items as a sequence of choices.

The *Plackett-Luce* model is then

`$$P(i_1 \succ \ldots \succ i_J) = \prod_{j=1}^J\frac{\alpha_{i_j}}{\sum_{i \in A_j} \alpha_i}$$`

where $A_j$ is the set of alternatives in choice $j$.

**PlackettLuce** can be used to fit this model.

---
@title[Neflix data]

## Example 1: Netflix Data

For the Netflix Prize, Netflix released several data sets 
comprising movie rankings.

Using `read.soc` from **PlackettLuce**, we read in a set of 
rankings for 4 movies

```{r}
library(PlackettLuce)
preflib <- "http://www.preflib.org/data/election/"
netflix <- read.soc(file.path(preflib,
                              "netflix/ED-00004-00000138.soc"))
head(netflix, 2)
```

+++

### Convert to Rankings

**PlackettLuce** requires the rankings to give the rank per item 
vs. item per rank.

We convert the rankings using `as.rankings`, which creates a 
special data structure

```{r}
R <- as.rankings(netflix[,-1], input = "ordering")
colnames(R) <- attr(netflix, "item")
print(R[1:3], width = 60)
```

---

### Fit Plackett-Luce Model

Now `PlackettLuce` can be used to fit the model, with frequencies 
as weights

```{r}
mod <- PlackettLuce(R, weights = netflix$n)
coef(mod, log = FALSE)
```

These coefficients are the *worth* parameters and represent the 
probability that each movie is ranked first.

---

### Inference

For inference it is better to work on the log scale. We must set one item as the reference (log-worth = 0).

We can use `qvcalc` to compute comparison intervals for all items

```{r qvcalc}
qv <- qvcalc(mod)
plot(qv, ylab = "Worth (log)", main = NULL)
```

```{r qvcalc-makeLink, results = "asis", echo = FALSE}
makeLink("qvcalc", title = "Worth of movies in Netflix data",
         alt = "Plot of estimated log-worth for each movie, with 95% comparison interval. Beverly Hills Cop is significantly more popular than the other three movies, Mean Girls is significant more popular than The Mummy Returns or Mission: Impossible II, but there was no significant difference in users’ preference for these last two movies.", width = 300)
```

---

## Ranking properties

The Netflix rankings are an example of *strict*, *complete*
rankings.

In other applications we might have
 * tied ranks
 * incomplete rankings
     - *sub-rankings*: only some items ranked each time
     - *top-n*: only the top $n$ items are ranked
     
**PlackettLuce** implements a generalized model which handles 
ties and sub-rankings.

---

## Generalized Model

A ranking is now an ordering of sets 

$$C_1 \succ C_2 \succ \ldots \succ C_J$$

and the generalized model with ties up to order $D$ is

`$$
\prod_{j = 1}^J \frac{f(C_j)}{
\sum_{k = 1}^{\min(D_j, D)} \sum_{S \in {A_j \choose k}} f(S)}
$$`

where

`$$f(S) = \delta_{|S|} \left(\prod_{i \in S} \alpha_i \right)^\frac{1}{|S|}$$`

---

## Ranking Networks

In some cases, the underlying network of wins and losses means 
the worth cannot be estimated by maximum likelihood.

@div[left-50]

```{r always-loses, message = FALSE, echo = FALSE, fig.width = 3.5, fig.height = 3.5, fig.show = "hold", out.width = "300px", echo = FALSE, results = "asis"}
par(mar = rep(1, 4))
library(igraph)
R <- matrix(c(1, 2, 0, 0,
              2, 0, 1, 0,
              1, 0, 0, 2,
              2, 1, 0, 0,
              0, 1, 2, 0), byrow = TRUE, ncol = 4,
            dimnames = list(NULL, LETTERS[1:4]))
R <- as.rankings(R)
A <- adjacency(R)
net <- graph_from_adjacency_matrix(A)
plot(net, edge.arrow.size = 0.5, vertex.size = 30, frame = TRUE)
makeLink("always loses", title = "Network in which one item always loses",
         alt = "Network in which one item always loses", width = 300)
```

@divend

@div[right-50]

```{r disconnected, message = FALSE, echo = FALSE, fig.width = 3.5, fig.height = 3.5, fig.show = "hold", out.width = "300px", echo = FALSE, results = "asis"}
par(mar = rep(1, 4))
R <- matrix(c(1, 2, 0, 0, 0, 
              2, 0, 1, 0, 0,
              0, 1, 2, 0, 0,
              0, 0, 0, 1, 2,
              0, 0, 0, 2, 1), byrow = TRUE, ncol = 5,
            dimnames = list(NULL, LETTERS[1:5]))
R <- as.rankings(R)
A <- adjacency(R)
net <- graph_from_adjacency_matrix(A)
plot(net, edge.arrow.size = 0.5, vertex.size = 30,
     frame = TRUE)
makeLink("disconnect", title = "Disconnected Network",
         alt = "Network with two separate groups of items, that are only observed to win or lose against other items in theor group", width = 300)
```



@divend

---

## Pseudo-rankings

**PlackettLuce** connects the network by adding `npseudo` 
*pseudo-rankings* with a ghost item.

@div[left-50]

@ul
- The MLE is always estimable
- Can be viewed as a Bayesian prior
- Default `nspeudo = 0.5`
@ulend

@divend

@div[right-50]

```{r pseudo-rankings, fig.width = 3.5, fig.height = 3.5, echo = FALSE, results = "asis"}
par(mar = rep(1, 4))
R <- matrix(c(0, 0, 0, 1, 2,
              0, 0, 0, 2, 1,
              0, 0, 1, 0, 2,
              0, 0, 2, 0, 1,
              0, 1, 0, 0, 2,
              0, 2, 0, 0, 1,
              1, 0, 0, 0, 2,
              2, 0, 0, 0, 1,
              1, 2, 0, 0, 0,
              2, 0, 1, 0, 0,
              1, 0, 0, 2, 0,
              2, 1, 0, 0, 0,
              0, 1, 2, 0, 0), byrow = TRUE, ncol = 5,
            dimnames = list(NULL, c(LETTERS[1:4], "G")))
R <- as.rankings(R)
A <- adjacency(R)
net <- graph_from_adjacency_matrix(A)
V(net)$color <- c(rep("#E69F00", 4), "#84aae8")
plot(net, edge.arrow.size = 0.5, vertex.size = 30, frame = TRUE)
makeLink("pseudo-rankings", title = "Network with pseudo-rankings",
         alt = "Network with pseudo-rankings, in which each item wins and loses against ghost item", width = 300)
```



@divend

